Depression and suicidality as credible signals of need

### 2021-05-22 Paper.rmd
Many minor edits, mostly to the bargaining theory section

Major depression and suicide are generally viewed as pathological responses to stress that elicit aversive responses from others. Alternatively, depression and suicide attempts might involve honest signals of need in response to stress that can increase support from reluctant others. To test this hypothesis, we examined responses to emotional signals in an experimental vignette study involving claims of substantial need in the presence of conflicts of interest and private information about whether the signaler was actually in need. In a sample of 1,240 participants recruited from Amazon Mechanical Turk, costlier signals like depression and suicide attempts resulted in greater perceptions of need, reduced perceptions of manipulativeness, and increased likelihood of support compared to simple verbal requests and crying without further symptoms. Additionally, as predicted, the effect of signal type on likelihood of support was largely mediated by the effect of signaling on belief in need. Our results support the hypothesis that depression and suicide attempts, apparent human universals, are credible signals of need.

@@

With a cooperative species like our own, the ubiquity of conflicts of interests means that there will always be disagreement over the levels of investment in a cooperative endeavor and the division of the resulting benefits. In many such cases, individuals who were dissatisfied with the terms of cooperation could seek other partners [@hammerstein2016]. In many other cases, however, it would be difficult or impossible to switch partners. Spouses who were dissatisfied with their partner's investment in their new infant, for example, would not easily find a different partner to invest in that infant [@Hagen:1999]. Similarly, an adolescent who was dissatisfied with her parent's investment in her would not easily find other parents who were willing to invest more, nor could parents easily produce another adolescent. In these examples, and many cooperative endeavors central to human biological fitness, each party has monopoly power over the benefits they bring to the endeavor -- no one is easily replaced [@Hagen2002; @Hagen2003]. Such interdependence is increasingly recognized to be important to the evolution of cooperation in humans and other animals [@roberts2005; @tomasello2012; @aktipis2018].

@@

Formal models of bargaining highlight the role of private information regarding the value each party places the cooperative endeavor. Withholding or risking one's contributions to a cooperative endeavor can be conceptualized as credibly signaling private information about one's need. Cooperation generates benefits over time. Thus, withholding or risking contributions incurs a *cost of delay* -- the more one values the endeavor, the greater the cost of delay. In a classic non-cooperative game theory model of bargaining over the division of benefits, @rubinstein1982 showed that two parties can come to an immediate agreement over division of benefits despite conflicts of interest if the parties' valuations of cooperation are not private information: in this case, each party knows exactly what division of benefits the other will accept, and can therefore make that offer immediately, avoiding the cost of delay. If valuations are private information, however, then immediate agreement might not be possible because each party has an incentive to request an amount greater than their actual valuation, which would harm the other party. A willingness to delay is a credible signal that one has a low valuation of the endeavor, and is therefore genuinely in need, because for those with a low valuation the benefit of waiting for a better offer outweighs the low cost of delay whereas for those for those with a high valuation the benefit of waiting for a better offer does not outweigh the higher cost of delay. Once valuations are known, the game reduces to the one in @rubinstein1982, and the parties can reach an agreement on the division of benefits [@KennanWilson1993; @feinberg2005]. @Hagen2003 proposed that the substantial reduction in productivity in depression corresponds to a willingness to delay, and is therefore a credible signal of low valuation and need. In other words, reduced productivity is relatively less costly for those whose efforts are currently not yielding many fitness-relevant benefits than for those whose efforts are yielding substantial benefits [@Hagen2003].

@@

Anthropology has long viewed suicidality as largely the result of social problems. Early in the field's history, anthropologists tackled the problem of suicide as manifested in the small, traditional societies in which they conducted their fieldwork. Suicide, they found, was commonly a form of protest, revenge, and/or appeal [@firth1936; @Firth:1961; @Malinowski:1932; @niehaus2012]. Some ethnographers conceptualized suicide as a form of anger or social pressure [@Hezel:1987; @Giddens:1964], whereas others emphasized the powerlessness of suicide victims [@Counts:1980].

@@

@@

In the bargaining framework, suicidality is conceptualized as putting all future fitness benefits at risk with some low but non-zero probability, credibly signaling low valuation of current circumstances [see @Rosenthal1993 for a similar game theory model of suicide attempts as costly signals]. On this view, most suicides deaths, especially in young, physically healthy individuals, would therefore be the inevitable consequence that some individuals lose this gamble. These and similar ideas [@andrews2006; @HagenEtAl2008; @nock2008; @Stengel1956] suggest that depression symptoms that put one's value to others at risk, such as loss of interest and suicidality, are credible signals of need.

Some evolutionary theorists have questioned the bargaining model based on evidence that depression is aversive to social partners [@nettle2004; @allen2006; @nesse2006; @myers2016]. These critics fail to acknowledge the central role of *conflict* in the signaling/bargaining models. Anger causes negative reactions in targeted social partners, yet it is probably an adaptation that exploits advantages in physical or social formidability to force concessions from others [@SellEtAl2009]. Aversive responses to bargaining are expected throughout the process, with potential helpers likely signaling the costliness of providing support in turn. Importantly, such responses do not prevent the delivery of meaningful benefits [@SymeHagen2021].

@@

```{r cdc, fig.width=8, fig.cap="US Suicidality non-fatal injury and death rates by age and sex (2001-2019)."}
library(hrafsuicidedata)
cdc2 %>% 
  na.omit() %>% 
  ggplot(aes(Age, `Crude Rate`, colour = Sex, linetype = Type)) + 
  geom_line() +
  hagenutils::scale_color_binary() +
  guides(
    colour = guide_legend(title=NULL),
    linetype = guide_legend(title=NULL, reverse = T)
  ) +
  labs(x = '\nAge (years)', y = 'Annual rate\n(per 100,000)') +
  theme_minimal(15) +
  theme(axis.title.y = element_text(angle = 0, hjust=0))
```

@@

The primary aim of this study was to examine responses to various forms of emotional signaling in an experimental vignette study. In the bargaining framework, the willingness to delay credibly signals the valuation of current cooperative arrangements, with greater delays indicating lower valuations. As signal cost increases, and signals therefore become less profitable to fake, we predicted that participants would report (1) increased belief in the signaler's claims and (2) increased likelihood of providing help, with (3) the increased likelihood of providing help mediated by the increased belief in the signaler's need. A secondary aim was to investigate a scenario that is common in traditional societies but uncommon in Western societies -- arranged marriages -- where conflicts between parents and offspring often result in depression and suicidality [@SymeEtAl2016]

@@

All participants were over 18, located in the United States or India, and had high quality MTurk metrics [completed at least 100 HITs with a HIT approval rate of over 98%; @kennedy2020]. Participants were excluded from the study if (1) they read the vignette too quickly (one-third of the time it took MG to read it), and (2) they failed clearly labeled attentions checks. The first attention check was shown immediately after the consent form and provided participants with a random word and asked them to enter the vowels in the order in which they are found in the word. The second attention check followed the vignette and involved asking three questions about the story that were easy to answer for anyone paying attention.

@@

All participants provided informed consent, and the consent form warned that some content might involve sexual assault. We estimated the study would take 4-8 minutes to complete for participants who did not take breaks. All participants who passed the attention checks were paid \$1 for their time, for an estimated rate of \$7.50/hr to \$15/hr (75% of US participants completed in `r signif(quantile(d_US$TotalTime, 0.75)/60, 2)` minutes; 75% of Indian participants completed in `r signif(quantile(d_India$TotalTime, 0.75)/60, 2)` minutes). This study was certified exempt by the Washington State University Human Research Protection Program.

@@

After rating their beliefs and actions, participants were randomized into either the control condition or one of four emotional signals by the victim (in order of increasing signal cost): (1) control condition: a verbal request without additional signaling; (2) crying; (3) mild-depression; (4) depression; and (5) a suicide attempt. The signals involve the participant encountering the victim some time after the adverse event and observing, e.g., crying, sad expressions, or poor personal hygiene. These descriptions do not use the terms depression, depressed, suicidal, or mental health.

@@

As both a validity check and a way to understand the degree participants would be willing to help if they believed the participant completely, we presented participants with strong evidence that the claims were true (*T3 Action*). In the US sample, this involved telling participants there was video evidence of the event in question occurring or a similar event after the fact. With the thwarted marriage vignette, this involved the participant seeing the man their daughter wants to marry trash talk their daughter and their family.

@@

According to our theoretical model, credible signals of need increase observers' belief that the victim is telling the truth. Increased belief then increases the likelihood of helping the victim (action). See Figure \@ref(fig:medmodel).

@@

The effects of the signals on rated beliefs and actions differed substantially by vignette, which we display in three ways for each signal in each vignette: estimated cumulative distributions plots, which show the entire distributions of rated beliefs and actions, including which signals and vignettes have high fractions of 0's and 1's (Figures \@ref(fig:ecdf)); regression estimates of the mean effects of the signals on beliefs and actions by vignette (Figure \@ref(fig:beliefvignette)); and between- and within-subjects signal effect sizes by vignette (Figures \@ref(fig:effectsizes) and \@ref(fig:effectsizes2)).

@@

The sociodemographic variables were strongly confounded with nationality (Indian participants were younger, lower income, with more years of education than US), which, in turn, were confounded with vignette (responses in the Indian Thwarted marriage vignette differed substantially from those in the US vignettes). We therefore conducted our exploration of the sociodemographic variables separately by nationality.

@@

These effects were largest in the "brother-in-law" and "romantic partner" vignettes, both of which involved claims of assault against participants' imagined daughters, and smaller in the "basketball coach" and the "thwarted marriage" vignettes. The small effect in the "basketball coach" vignette, might have been due to the lack of relatedness between participants and the victim (the star player), or how participants weighted the costs of suspending the coach vs. punishing a potentially innocent person.

The small signal effect in the "thwarted marriage" vignette, which involved the Indian sample, could indicate that our results do not generalize across cultures, undermining our adaptationist hypothesis. It could also reflect our poor understanding of contemporary Indian culture regarding dowry (the effect was larger in older participants). Baseline belief in the older daughter, and likelihood of helping her, was relatively high at baseline (greater than 50%) compared to victims in the other vignettes. Private information and conflict therefore probably played a smaller role and thus costly signals were less necessary. We observed a similar pattern in our pilot study, in which baseline belief in the victim's need was high, and costly signals had smaller effects than they did in the current study. Additionally, supporting the older daughter came at the cost of one's younger daughter, which may also help explain the relatively small signal effects. A final consideration is that data from the Indian sample appeared to be of lower quality, limiting our confidence in any of these interpretations (see limitations for more information).

@@

>>>>

Major depression and suicide are generally viewed as pathological responses to stress that elicit aversive responses from others. Alternatively, depression and suicide attempts might involve honest signals of need in response to stress that can increase support from reluctant others. To test this hypothesis, we examined responses to emotional signals in an experimental vignette study involving claims of substantial need in the presence of conflicts of interest and private information about whether the signaler was actually in need. In a sample of 1,240 participants recruited from Amazon Mechanical Turk, costlier signals like depression and suicide attempts resulted in greater perceptions of need, reduced perceptions of manipulativeness, and increased likelihood of support compared to simple verbal requests and crying without further symptoms. Additionally, as predicted, the effect of signaling on likelihood of support was largely mediated by the effect of signaling on belief in need. Our results support the hypothesis that depression and suicide attempts, apparent human universals, are credible signals of need.

@@

With a cooperative species like our own, ubiquitous conflicts of interest means that there will always be disagreement over the levels of investment in a cooperative endeavor and the division of the resulting benefits. In many such cases, individuals who were dissatisfied with the terms of cooperation could seek other partners [@hammerstein2016]. In many other cases, however, it would be difficult or impossible to switch partners. Spouses who were dissatisfied with their partner's investment in their new infant, for example, would not easily find a different partner to invest in that infant [@Hagen:1999]. Similarly, an adolescent who was dissatisfied with her parent's investment in her would not easily find other parents who were willing to invest more, nor could parents easily produce another adolescent. In these examples, and many cooperative endeavors central to human biological fitness, all parties have monopoly power over the benefits they bring to the endeavor -- no one is easily replaced [@Hagen2002; @Hagen2003]. Such interdependence is increasingly recognized to be important to the evolution of cooperation in humans and other animals [@roberts2005; @tomasello2012; @aktipis2018].

@@

Bargaining models assume that delaying cooperation is costly so that there is an incentive to quickly agree on a division of benefits, and more so for those who highly value the fruits of the cooperative endeavor. In a classic non-cooperative game theory model of bargaining, @rubinstein1982 showed that two parties can come to an immediate agreement over division of benefits despite conflicts of interest if the parties' valuations of cooperation are not private information: in this case, each party knows exactly what division of benefits the other will accept, and can therefore make that offer immediately, avoiding the cost of delay.

If valuations are private information, however, then costly delays might be unavoidable because each party has an incentive to request more than their actual valuation, and to reject the likely inflated requests from partners, leading to multiple rounds of bargaining. Models of bargaining with private information have a close relationship to models of credible signaling. A willingness to delay (i.e., refuse offers) credibly reveals one's low valuation of the endeavor, and therefore genuine need -- the benefit of waiting for a better offer outweighs the low cost of delay. Eagerness to reach a deal, on the other hand, credibly reveals a high valuation -- the benefit of waiting for a better offer does not outweigh the higher cost of delay. Once valuations are known, the game reduces to the one in @rubinstein1982, and the parties can reach an agreement on the division of benefits [@KennanWilson1993]. Delays also typically require that additional factors come into play [@feinberg2005 and references therein].

@Hagen2003 proposed that the substantial reduction in productivity in depression corresponds to a willingness to delay, and is therefore a credible signal of low valuation and need. In other words, reduced productivity is relatively less costly for those whose efforts are currently not yielding many fitness benefits than for those whose efforts are yielding substantial benefits.

<!-- Formal models of bargaining highlight the role of private information regarding the value each party places the cooperative endeavor. Withholding or risking one's contributions to a cooperative endeavor can be conceptualized as credibly signaling private information about one's need. Cooperation generates benefits over time. Thus, withholding or risking contributions incurs a *cost of delay* -- the more one values the endeavor, the greater the cost of delay.  -->

@@

Anthropology has long viewed suicidality as largely the result of social problems. Early in the field's history, anthropologists tackled the problem of suicide as manifested in the small, traditional societies in which they conducted their fieldwork. Suicide, they found, was commonly a form of protest, revenge, and/or appeal [@firth1936; @Firth:1961; @Malinowski:1932; @niehaus2012]. Some ethnographers emphasized suicide as a form of anger or social pressure [@Hezel:1987; @Giddens:1964], whereas others emphasized the powerlessness of suicide victims [@Counts:1980].

@@

```{r cdc, fig.width=8, fig.cap="US Suicidality non-fatal injury and death rates by age and sex (2001-2019). Data from @CDC2021."}
library(hrafsuicidedata)
cdc2 %>% 
  na.omit() %>% 
  ggplot(aes(Age, `Crude Rate`, colour = Sex, linetype = Type)) + 
  geom_line() +
  hagenutils::scale_color_binary() +
  guides(
    colour = guide_legend(title=NULL),
    linetype = guide_legend(title=NULL, reverse = T)
  ) +
  labs(x = '\nAge (years)', y = 'Annual rate\n(per 100,000)') +
  theme_minimal(15) +
  theme(axis.title.y = element_text(angle = 0, hjust=0))
```

@@

In the bargaining framework, suicidality is conceptualized as putting all future contributions to cooperative endeavors with social partners at risk with some low but non-zero probability, credibly signaling low valuation of current circumstances [see @Rosenthal1993 for a game theory model of suicide attempts as costly signals]. On this view, most suicides deaths, especially in young, physically healthy individuals, would therefore be the inevitable consequence that some individuals lose this gamble. These and similar ideas [@andrews2006; @HagenEtAl2008; @nock2008; @Stengel1956] suggest that depression symptoms that put one's value to others at risk, such as loss of interest and suicidality, are credible signals of need.

Some evolutionary theorists have questioned the bargaining model based on evidence that depression is aversive to social partners [@nettle2004; @allen2006; @nesse2006; @myers2016]. These critics failed to recognize the central role of *conflict* in the signaling/bargaining models. Anger is aversive to targeted social partners, yet it is probably an adaptation that exploits advantages in physical or social formidability to force concessions from them [@SellEtAl2009]. Aversive responses to depressive bargaining are likewise expected throughout the process, with potential helpers signaling the costliness of providing support in turn. Importantly, as with anger, aversive responses to depression do not prevent concessions by interdependent social partners [@SymeHagen2021].

@@

@@

The primary aim of this study was to examine responses to emotional signals that varied from low to high cost, in an experimental vignette study in which a possible victim of adversity asks for help from the participant, but has incentives to exaggerate her need. In the bargaining framework, the willingness to delay credibly signals the valuation of current cooperative arrangements, with greater delays indicating lower valuations. As signal cost increased, we predicted that participants would report (1) increased belief in the signaler's claims and (2) increased likelihood of providing help, with (3) the increased likelihood of providing help mediated by the increased belief in the signaler's need.

<!-- A secondary aim was to investigate a scenario that is common in traditional societies but uncommon in Western societies -- arranged marriages -- where conflicts between parents and offspring often result in depression and suicidality [@SymeEtAl2016] -->

@@

All participants were over 18, located in the United States or India, and had high quality MTurk metrics [completed at least 100 HITs with a HIT approval rate of over 98%, @kennedy2020]. Participants were excluded from the study if (1) they read the vignette too quickly (one-third of the time it took MG to read it), and (2) they failed clearly labeled attentions checks. The first attention check was shown immediately after the consent form and provided participants with a random word and asked them to enter the vowels in the order in which they are found in the word. The second attention check followed the vignette and involved asking three questions about the story that were easy to answer for anyone paying attention.

@@

All participants provided informed consent, and the consent form warned that some content might involve sexual assault. We estimated the study would take 4-8 minutes to complete for participants who did not take breaks [MTurkers commonly multitask, or leave the survey page and return later, @NeckaEtAl2016]. All participants who passed the attention checks were paid \$1 for their time, for an estimated rate of \$7.50/hr to \$15/hr (75% of US participants completed in `r signif(quantile(d_US$TotalTime, 0.75)/60, 2)` minutes; 75% of Indian participants completed in `r signif(quantile(d_India$TotalTime, 0.75)/60, 2)` minutes). This study was certified exempt by the Washington State University Human Research Protection Program.

@@

After rating their beliefs and actions, and which emotions they thought the potential victim was experiencing, participants were randomized into either the control condition or one of four emotional signals by the victim (in order of increasing signal cost): (1) control condition: a verbal request without additional signaling; (2) crying; (3) mild-depression; (4) depression; and (5) a suicide attempt. The signals involved the participant encountering the victim some time after the adverse event and observing, e.g., crying, sad expressions, or poor personal hygiene. These descriptions did not use the terms depression, depressed, suicidal, or mental health.

@@

As both a validity check and a way to understand the degree participants would be willing to help if they believed the participant completely, we presented participants with strong evidence that the claims were true. In the US sample, this involved telling participants there was video evidence of the event in question occurring or a similar event after the fact. With the thwarted marriage vignette, this involved the participant seeing the man their daughter wants to marry trash talk their daughter and their family. Participants were then asked to rate their likelihood of acting (*T3 Action*).

@@

According to our theoretical model, emotional signals that put cooperative benefits at risk increase observers' belief that the victim is genuinely in need. Increased belief then increases the likelihood of helping the victim (action). See Figure \@ref(fig:medmodel).

@@

The effects of the signals on rated beliefs and actions differed substantially by vignette, which we display in three ways for each signal in each vignette: estimated cumulative distributions plots, which show the entire distributions of rated beliefs and actions, including which conditions have high fractions of 0's and 1's (Figures \@ref(fig:ecdf)); regression estimates of the mean effects of the signals on beliefs and actions by vignette (Figure \@ref(fig:beliefvignette)); and between- and within-subjects signal effect sizes by vignette (Figures \@ref(fig:effectsizes) and \@ref(fig:effectsizes2)).

@@

The sociodemographic variables were strongly confounded with nationality (Indian participants were younger, lower income, with more years of education than US; see Figure \@ref(fig:socioeconomic)), which, in turn, were confounded with vignette (responses in the Indian Thwarted marriage vignette differed substantially from those in the US vignettes). We therefore conducted our exploration of the sociodemographic variables separately by nationality.

@@

These effects were largest in the "brother-in-law" and "romantic partner" vignettes, both of which involved claims of assault against participants' imagined daughters, and smaller in the "basketball coach" and the "thwarted marriage" vignettes. The small effect in the "basketball coach" vignette might have been due to the lack of relatedness between participants and the victim (the star player), or how participants weighted the costs of suspending the coach vs. punishing a potentially innocent person.

The small signal effect in the "thwarted marriage" vignette, which involved the Indian sample, could indicate that our results do not generalize across cultures, undermining our adaptationist hypothesis. It could also reflect our poor understanding of contemporary Indian culture regarding dowry (the effect was larger in older participants). Baseline belief in the older daughter, and likelihood of helping her, was relatively high at baseline (`r signif(100*mean(d$T1Belief[d$vignette=='Thwarted marriage']), 2)`%) compared to victims in the other vignettes. Private information and conflict therefore probably played a smaller role and thus costly signals were less necessary. We observed a similar pattern in our pilot study, in which baseline belief in the victim's need was high, and costly signals had smaller effects than they did in the current study. Additionally, supporting the older daughter came at the cost of one's younger daughter, which may also help explain the relatively small signal effects. A final consideration is that data from the Indian sample appeared to be of lower quality, limiting our confidence in any of these interpretations (see limitations for more information).

@@

### Measureing Beliefs and Actions
